Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-23 ind_files/06_nontechnical_discussion.tex
Minor rephrasing

The equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection' (following previous studies such as~\cite{constable_demographic_2016} and~\cite{week_white_2021}). Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models, especially in social evolution, illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010,melbinger_evolutionary_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,debarre_evolutionary_2016, behar_fluctuations-induced_2016, constable_demographic_2016,veller_drift-induced_2017,abu_awad_effects_2018,parsons_pathogen_2018,mcavoy_public_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Chapter \ref{chap_unification} explains how the  results of all these studies can be explained using noise-induced selection. Similar effects have been recognized to play a role in specific finite population models of heterogamety~\citep{veller_drift-induced_2017,saunders_sex_2018}, life-history evolution~\citep{gillespie_natural_1974,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019,wang_reproductive_2023}, and epidemiology~\citep{kogan_two-strain_2014,parsons_pathogen_2018,day_price_2020}.

This thesis is, to the best of my knowledge, the first time that noise-induced selection has been explicitly formulated and derived in full generality using a general `model-independent' evolutionary framework such as the Price equation (or my generalization thereof). My results thus unify several previous studies focused on specific systems under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations of any nature as long as there is some heritable variation in per-capita turnover rates. Most recently, a study showed (among other things) that in a very broad class of competition models of population ecology, finite populations can directionally evolve in a manner opposite to the direction of evolution predicted in the infinite population limit, essentially due to the effects of noise-induced selection~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\citet{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a form that highlights connections with standard results in evolutionary theory such as the Price equation and Fisher's fundamental theorem (see section \ref{sec_fun_theorems}). It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

>>>>

Several specific finite population models illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010,melbinger_evolutionary_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,debarre_evolutionary_2016, behar_fluctuations-induced_2016, constable_demographic_2016,veller_drift-induced_2017,abu_awad_effects_2018,parsons_pathogen_2018,mcavoy_public_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic evolution'~\citep{constable_demographic_2016,mcleod_social_2019, wang_reproductive_2023}. These effects have been recognized to play a role in specific finite population models of heterogamety~\citep{veller_drift-induced_2017,saunders_sex_2018}, life-history evolution~\citep{gillespie_natural_1974,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019,wang_reproductive_2023}, and epidemiology~\citep{kogan_two-strain_2014,parsons_pathogen_2018,day_price_2020}.

As explained in chapter \ref{chap_unification}, my general birth-death formalism reveals that the results of all these studies can be explained as specific instances of a more general phenomenon we call `noise-induced selection' (following previous studies such as~\cite{constable_demographic_2016} and~\cite{week_white_2021}). In particular, the equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force --- noise-induced selection --- acting on variation in per-capita turnover rate $\tau$ and favoring types whose per-capita turnover rate is lower than that of the population average. Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. This thesis is, to the best of my knowledge, the first time that noise-induced selection has been explicitly formulated and derived in full generality using a general `model-independent' evolutionary framework such as the Price equation (or my generalization thereof). My results thus unify several previous studies focused on specific systems under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations of any nature as long as there is some heritable variation in per-capita turnover rates.

A recent study showed (among other things) that in a very broad class of competition models of population ecology, finite populations can directionally evolve in a manner opposite to the direction of evolution predicted in the infinite population limit, essentially due to the effects of noise-induced selection~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\citet{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a form that highlights connections with standard results in evolutionary theory such as the Price equation and Fisher's fundamental theorem (see section \ref{sec_fun_theorems}). It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.
